An efficient acyclic contact planner for multiped robots

### 2016-01-26 body/conclusion.tex
finished pass with andrea

In this paper we consider acyclic contact planning in cluttered environments, formulated as two sub problems that we address sequentially.
The first problem is \Pa, that consists in computing a guide path for the root of the robot that can be extended; the second problem is $\mathcal{P}_2$, that is the computation of a discrete sequence of contacts along the root path.
Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. We introduced a low-dimensional space $C_{reach}$ that can be mapped 
into the contact sub-manifold of the robot, approximated, and efficiently sampled by our Reachability-Based planner.
Our contribution to \Pb is a pragmatic contact generation scheme that can take into
account criteria to enforce interesting properties on the generated contacts (such as robustness, energy efficiency or naturalness).

Aside from the theoretical contributions, our results demonstrate that our method allows a very interesting compromise between three 
criteria that are hard to conciliate: generality, performance, and quality of the solution, making it the first acyclic contact
planner compatible with \gls{interactive} applications.
%
\textbf{Regarding generality}, the \gls{reachc}, coupled with an approach based on limb decomposition, 
allows the method to address arbitrary multiped robots. The only pre-requisite is the specification 
of the volumes $W^0$.
%
\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and $\mathcal{P}_2$. This results in computation costs close to real-time in
known static environments.
%
\textbf{Regarding the quality of the paths}, a parametrization of the \gls{reachc} allows us to compute
relevant paths in all the presented scenarios, with low rejection rates.
As for \cite{Bouyarmane2009}, failures can still occur, due to the compromise criterion used in computing the guide path.
The low computational burden of our framework however allows for fast re-planning in case of failure.

One direction for future work is to focus on a more accurate formulation of $C_{reach}$ to address 
this limitation.
Our main objective is of course to generate the complete interpolation between the contact sequences.
We have already obtained some successes for some of the computed sequences~\citep{Carpentier2016}, but we admit that additional
work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating
conditions that guarantee that the interpolation between two contact configurations is dynamically feasible.
Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
at the center of mass of the robot.
 We are also
working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
Lastly, we aim at going beyond static equilibrium and to perform kinodynamic planning in the configuration space. We believe that the most promising direction in this regard is to integrate
the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
Addressing these three issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.

>>>>

In this paper we consider the \gls{cluttered} contact planning problem, formulated as two sub-problems that we address sequentially.
The first problem $\mathcal{P}_1$ consists in computing an \gls{equilibrium feasible} guide path for the root of the robot;
 the second problem $\mathcal{P}_2$ is the computation of a discrete sequence of whole-body configurations along the root path.

Our contribution to \Pa is the introduction of a low-dimensional space $C_{reach}$, an approximation of the space of \gls{equilibrium feasible} root configurations.
Thanks to the computationally efficient verification of the \gls{reachc}, we are able to solve \Pa much faster than previous approaches.

%~ Our contribution to \Pa is a generic characterization of the properties that the guide path must satisfy, in particular to enforce the completeness of the acyclic contact planner. 
Our contribution to \Pb is a fast contact generation scheme that can take into
account criteria to optimize user-defined properties.

Our results demonstrate that our method allows a pragmatic compromise between three 
criteria that are hard to conciliate: generality, performance, and quality of the solution, making it the first acyclic contact
planner compatible with \gls{interactive} applications.
%
\textbf{Regarding generality}, the \gls{reachc}, coupled with an approach based on limb decomposition, 
allows the method to address arbitrary multiped robots in \gls{cluttered} problems. The only pre-requisite is the specification 
of the volumes $W^0$.
%
\textbf{Regarding performance}, our framework is really efficient in addressing both \Pa and $\mathcal{P}_2$. This results in \gls{interactive} computation times.
%
\textbf{Regarding the quality of the paths}, the \gls{reachc} allows us to compute
\gls{equilibrium feasible} paths in all the presented scenarios, with low rejection rates.
As for \cite{Bouyarmane2009}, failures can still occur, due to the approximate condition used to compute the guide path.
The low computational burden of our framework however allows for fast re-planning in case of failure.

One direction for future work is to focus on a more accurate formulation of $C_{reach}^0$ to address 
this limitation.
Our main objective is of course to generate the complete interpolation between the contact sequences.
We have already obtained some success for some of the computed sequences~\citep{Carpentier2016}, but additional
work is required on the planning side to obtain a seamless workflow. To achieve this, we are currently working on the notion of transition certificate, i.e. formulating
conditions that guarantee that the interpolation between two contact configurations is dynamically feasible.
%~ Our current direction is to propose a geometric quantification of both the kinematic and dynamic constraints of the problem, expressed
%~ at the center of mass of the robot.
 %~ We are also
%~ working on new heuristics more closely related to the dynamics of the systems (for instance, choose the contacts that minimize the jerk between two contact configurations).
Lastly, we aim at going beyond static equilibrium and performing kinodynamic planning. We believe that the most promising direction in this regard is to integrate
the notion of Admissible Velocity Propagation in our current work \citep{DBLP:conf/rss/PhamCN13}.
Addressing these two issues is essential to bridge the gap between the planning and control aspects of multiped locomotion.

### 2016-01-26 body/contact.tex
complete, need abstract and section 4
added missing figure on feasibility
finished pass with andrea

% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
Our planner computes guide paths in \gls{creach0}, an over-approximation of the set of \gls{equilibrium feasible} paths, \gls{ccequil0}.
As an input of this stage, we however assume an \gls{equilibrium feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow \mathbb{R}^r$ for the root of the robot $R$.
If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section TODO.
We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.

In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a given limb 
and root location.
Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.

@@

\subsection{Definition of a contact generator}
\label{sec:single_contact}
We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$.
From \glslink{contact feasible}{\textit{contact feasibility}} we know that there exists at least one $\mathbf{q}^{\overline{0}}$ that verifies $\mathbf{q}^k \oplus \mathbf{q}^{\overline{k}} \in$ \gls{ccontact}.
The issue is then to exhibit locally this $\mathbf{q}^{\overline{0}}$.

While exhibiting analytically a $\mathbf{q}^{\overline{0}}$ does not seem tractable, we can iteratively try to generate one as follows:
\begin{enumerate}
\item Generate randomly a collision-free limb configuration;
\item Project the end-effector onto the closest surface with inverse kinematics;
\item If a valid solution is found, stop. Otherwise repeat from step 1.
\end{enumerate}

We partially trade the completeness property of this approach for a more efficient solution introduced in our previous work~\citep{Tonneau2014}, which we describe
here for completeness.

We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimal distance 
between an effector and a surface of the environment is less than a small value $\epsilon$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
\item According to a user-defined heuristic $h$, sort the elements of $S$;
\item Select the first configuration of $S$. Project the configuration onto contact with inverse kinematics. If $S$ is empty, stop (failure case) (Figure~\ref{fig:contact_gen} (d) and (e));
\item If a valid solution is found, stop (success case). Otherwise remove the element from $S$ and go back step 5.
\end{enumerate}

@@

It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
the failure to generate a contact.

We thus have defined a real-time contact generator. For a given configuration, the generator can be as complete as required by the context (by increasing the number of samples).

@@

As an input, we consider the guide path $\mathbf{q}^0(t)$, automatically discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations. To ensure continuity in the contact transition phases, we propose a recursive mapping $\pi$, for any $0<i<j$:
\begin{equation*}
    \pi\colon\left\{
    \begin{aligned}		
        \mathbf{Q}^0 \in C_{equil}^0 & \longrightarrow C_{equil} \\
        %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
        \mathbf{q}^{0}_i &  \longrightarrow  \mathbf{q}^{0}_i \oplus g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
    \end{aligned}
    \right.
\end{equation*} 
$g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.

@@

Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
If possible, a limb in contact at time $i-1$ remains in contact at time $i$. The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. The solver is directly provided by the HPP software.

If the solver fails, the contact is broken and a collision-free configuration is assigned to the limb.
If two or more contacts are broken in a single time step, intermediate steps are added to reposition the contacts one at a time, before moving.

@@

%\subsection{Generation of a contact configuration}  
\subsection{Creating contacts.}
Contacts are created using three rules:
\begin{enumerate}
\item Only one contact creation can happen between two consecutive states. This increases the odds that the interpolation between the two states be feasible;
%\adnote{Before you said that you ensured that transition between two states is feasible...}
\item A contact is validated if and only if the resulting configuration is in static equilibrium, no matter what user-defined heuristics are used; 
\item We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others have been treated.
\end{enumerate}

These three rules provide a deterministic manner to generate contacts along the discretized guide path.
For a given \gls{equilibrium feasible} path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}, thanks to 
the heuristics that we use for contact selection.

>>>>

% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
Our planner computes guide paths in \gls{creach0}, an approximation of \gls{ccequil0}.
As an input of this stage, we however assume an \gls{equilibrium feasible} root guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow$ \gls{ccequil0}.
If this is not the case, our planner will fail rapidly, thus allowing replanning, as discussed in Section~\ref{sec:perf}.
We now consider the second problem of computing a discrete sequence of equilibrium configurations $\mathbf{Q}^{\overline{0}}$ along $\mathbf{q}^0(t)$.

In this Section we first describe a single contact-generation process, that is how to generate a contact configuration for a limb, given a
root location.
Then, we propose an iterative algorithm to generate a discrete sequence of contact configurations in static equilibrium.

@@

\subsection{Definition of a contact generator}
\label{sec:single_contact}
Given a configuration $\mathbf{q}^{\overline{k}}$ of the root and all the limbs but $R^k$, we look for a limb configuration $\mathbf{q}^k$ such that
$R^k$ is in contact, and not colliding (neither with parts of the robot nor with the environment).
While exhibiting analytically a $\mathbf{q}^{k}$ does not seem tractable, we can iteratively try to generate one as follows:
\begin{enumerate}
\item Generate randomly a collision-free limb configuration;
\item Project the end-effector onto the closest surface with inverse kinematics;
\item If a valid solution is found, stop. Otherwise repeat from step 1.
\end{enumerate}

We trade the completeness of this approach for a more efficient solution introduced in our previous work~\citep{Tonneau2014}, which we describe
here to be exhaustive.

We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimum distance 
between an effector and a surface of the environment is less than a small value $\epsilon$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree data structure;
\item At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact (Figure~\ref{fig:contact_gen} (b) and (c));
\item According to a user-defined heuristic $h$, sort the elements of $S$;
\item Select the first configuration of $S$. Project the configuration onto contact with inverse kinematics. If $S$ is empty, stop (failure case) (Figure~\ref{fig:contact_gen} (d) and (e));
\item If a valid solution is found, stop (success case). Otherwise remove the element from $S$ and go back to step 5.
\end{enumerate}

@@

%~ It should be noted that a failure case for our contact generator does not necessarily mean that the planner will globally fail. Indeed,
%~ for a given root configuration, if contact configurations are found for other limbs, resulting in static equilibrium, the configuration will be accepted despite
%~ the failure to generate a contact.

@@

As an input, we consider the guide path $\mathbf{q}^0(t)$, discretized into a sequence of $j$ key configurations:  
\begin{equation*}
	\mathbf{Q}^0 = [\mathbf{q}^0_{0}; \mathbf{q}^0_{i}; ..., \mathbf{q}^0_{j-1}]
\end{equation*} 
where $\mathbf{q}^0_{0}$ and $\mathbf{q}^0_{j-1}$ respectively correspond to the start and goal configurations. %To ensure continuity in the contact transition phases,
We want to extend the configurations of $\mathbf{Q}^0$ in such a way that continuity is preserved regarding the contact transitions.
To do so, we define an algorithm that, given the current root configuration, and the previous full body configuration, computes a full body configuration 
in \gls{ccequil} such that contacts are maintained if possible.
The first full body configuration of the sequence is given by the initial state of the robot.

 %~ we propose a recursive mapping $\pi$, for any $0<i<j$:
%~ \begin{equation*}
    %~ \pi\colon\left\{
    %~ \begin{aligned}		
        %~ \mathbf{Q}^0 \in C_{Equil}^0 & \longrightarrow C_{Equil} \\
        %~ %\mathbf{q}^{0}_0 &  \longrightarrow  \mathbf{q}_{start} \\
        %~ \mathbf{q}^{0}_i &  \longrightarrow  g(\mathbf{q}_{i - 1},\mathbf{q}^{0}_i) 
    %~ \end{aligned}
    %~ \right.
%~ \end{equation*} 
%~ $g$ is the method that extends a root configuration into a full-body configuration. At each step, it tries to generate a contact configuration that preserves as much as possible
%~ the previous contacts while allowing for static equilibrium. The objective is thus to characterize $g$.
%~ We initialize the recurrence with $\pi(\mathbf{q}^0_{0}) = \mathbf{q}_0$ the initial configuration of the robot.

@@

Figure~\ref{fig:break_contact} illustrates the contact-persistence strategy.
If possible, a limb in contact at step $i-1$ remains in contact at step $i$. The contact is broken if an inverse-kinematics solver fails to find a collision-free limb configuration that satisfies joint limits. The solver is directly provided by the Hpp software.

If the solver fails, the contact is broken and a collision-free configuration is assigned to the limb.
If two or more contacts are broken in a single step, one or more intermediate steps are added.
For these steps the root configuration is the same as for the previous step, with the difference that
one faulty contact is repositioned, in the hope that it will not be broken at the next step.

@@

%\subsection{Generation of a contact configuration}  
\subsubsection{Creating contacts.}
Contacts are created using three rules:
\begin{enumerate}
\item Only one contact creation can happen between two consecutive steps. This increases the odds that the interpolation between the two steps be feasible;
%\adnote{Before you said that you ensured that transition between two states is feasible...}
\item A contact is validated if and only if the resulting configuration is in static equilibrium; 
\item We use a FIFO approach:  we always try first to create a contact with the limb that has been contact-free the longest. If the contact creation
was not successful for a limb, the limb is pushed on top of the queue, and will only be tried again after the others.
\end{enumerate}

These three rules provide a deterministic generation of contacts along the discretized guide path.
For a given \gls{equilibrium feasible} path, our current implementation does not guarantee that the planner will succeed, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}.

### 2016-01-26 body/efort.tex
finished pass with andrea

% !TEX root =  ../main.tex
\subsection{Manipulability-based heuristics for locally-optimal contact selection}
The other kind of heuristics that we use are based on the analysis of the motion capabilities
provided by a candidate configuration regarding the task being achieved.
These heuristics allow to locally optimize a contact location with respect to a particular task.
For instance, one can be interested in configurations that allow to efficiently exert a force compatible with the global direction of motion,
a high velocity in a given direction, or to stay away from singular configurations.
In this section, we derive three such heuristics from the work on manipulability by \cite{Yoshikawa1984}, that we present first.

\subsubsection{The force and velocity ellipsoids.}
For a limb configuration $\mathbf{q}^k$, the Jacobian matrix
 $\mathbf{J}(\mathbf{q}^k)$   defines the relation:

\begin{equation} \label{jac}
\mathbf{\dot{p}^k} = \mathbf{J}(\mathbf{q}^k) \dot{\mathbf{q}}^k
\end{equation}

For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}(\mathbf{q}^k)$ as $\mathbf{J}$.

As a linear approximation of the forward-kinematics function $f$ , $\mathbf{J}$ describes how small
variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
It can be seen as the transformation mapping a velocity in the configuration space into a velocity in
the Euclidian space.

@@

We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations that we discard).
From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
\begin{equation} \label{jaceq}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
\end{equation}

@@

\begin{equation} \label{ellipsoid}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1
\end{equation}
This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of velocities that can
be reached under the constraint \eqref{ball} for the current configuration.
The longer an axis of the ellipsoid is, the faster the configuration can move along the direction of the axis.
Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.

\begin{figure}[htb]
  \centering
	\begin{overpic}[width=1\linewidth]{figures/EFORT/ellipsoid}
		\put (8.5,1.2) {\small{Velocity ellipsoid}}
		\put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
	\end{overpic}
  % \includegraphics[width=1\linewidth]{ressources/screens/sampling/canap/bd2_cmyk}
  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 dofs and 2 segments.
Only the horizontal and vertical speeds are shown (not the rotation speeds), since it would require being able to draw in four dimensions.}
		   \label{sec:efort_ellipsoid}
\end{figure}

@@

\begin{equation}
h_{EFORT}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu_0 \mathbf{n}^T \mathbf{m})
\end{equation}
where $\mu_0$ and $\mathbf{n}$ are respectively the friction coefficient and the normal vector of the contact surface.

If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
\begin{equation}
h_{vel}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{m}]^{-\frac{1}{2}} ( \mu_0 \mathbf{n}^T \mathbf{m})
\end{equation}

$h_{EFORT}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
In other less demanding cases, manipulability can also be considered to avoid singularities.
To do so, we can consider the manipulability measure $w$, also given by \citeauthor{Yoshikawa1984}:

\begin{equation*} \label{ellipsoid}
w(\mathbf{q}) = \sqrt{det(\mathbf{J}\mathbf{J}^T)}
\end{equation*}
$w$ measures the ``distance'' between a given configuration and a singular configuration. When $w$ is equal to 0, the configuration is in a singular state;
the greater $w$ is, the further away the configuration is from singularity.
We then derive:
\begin{equation}
h_{w}(\mathbf{q}, \mathbf{m}) = w(\mathbf{q})  (\mu_0 \mathbf{n}^T \mathbf{m})
\end{equation}

$h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
can be precomputed off-line), that allow the planner to select the best
candidates with respect to the local motion task along the guide path.
This results in faster computation times, and a reduced number of overall contact switches.

>>>>

% !TEX root =  ../main.tex
\subsection{Manipulability-based heuristics for contact selection}
This Section proposes heuristics to select a contact that optimizes desiered capabilities.
%~ These heuristics allow to locally optimize a contact location with respect to a particular task.
For instance, one can be interested in configurations that allow to efficiently exert a force in the global direction of motion,
a high velocity in a given direction, or to stay away from singular configurations.
In this section, we derive three such heuristics from the work on manipulability by \cite{Yoshikawa1984}, that we present first.

\subsubsection{The force and velocity ellipsoids.}
For a limb configuration $\mathbf{q}^k$, the Jacobian matrix
 $\mathbf{J}^k(\mathbf{q}^k)$   defines the relation:

\begin{equation} \label{jac}
\mathbf{\dot{p}}^k = \mathbf{J}^k(\mathbf{q}^k) \dot{\mathbf{q}}^k
\end{equation}

For clarity in the rest of the section we omit the $k$ indices and write $\mathbf{J}^k(\mathbf{q}^k)$ as $\mathbf{J}$.

As a linear approximation of a forward-kinematics function, $\mathbf{J}$ describes how small
variations from the configuration $\mathbf{q}$ affect the position vector $\mathbf{p}$.
%~ It can be seen as the transformation mapping a velocity in the configuration space into a velocity in
%~ the Euclidian space.

@@

We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations, which we discard).
From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
\begin{equation} \label{jaceq}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
\end{equation}

@@

\begin{equation} \label{ellipsoid}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} \leq 1
\end{equation}
This ellipsoid is called the manipulability ellipsoid, or velocity ellipsoid, introduced by \cite{Yoshikawa1984}. It describes the set of end-effector velocities that can
be reached under the constraint \eqref{ball} for the current configuration.
The longer the axis of the ellipsoid is, the faster the end-effector can move along the direction of the axis.
Figure~\ref{sec:efort_ellipsoid} - left shows the velocity ellipsoid for different configurations of a manipulator with two degrees of freedom.

\begin{figure}[!tbp]
  \centering
	\begin{overpic}[width=1\linewidth]{figures/EFORT/ellipsoid}
		\put (8.5,1.2) {\small{Velocity ellipsoid}}
		\put (62.5,1.2) {\small{Force ellipsoid} \tiny{(scale 0.5)}}
	\end{overpic}
  % \includegraphics[width=1\linewidth]{ressources/screens/sampling/canap/bd2_cmyk}
  \caption{Examples of velocity and force ellipsoids for a manipulator composed of 2 dofs and 2 segments.
Only the horizontal and vertical speeds are shown (not the rotation speeds), since it would require being able to draw in four dimensions.}
		   \label{sec:efort_ellipsoid}
\end{figure}

@@

\begin{equation}
h_{EFORT}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)\mathbf{m}]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \mathbf{m})
\end{equation}
where $\mu$ and $\mathbf{n}$ are respectively the friction coefficient and the normal vector of the contact surface.

If the ability to generate large velocities at the effector is considered, we define a new heuristic $h_{vel}$ with a similar reasoning on the velocity ellipsoid:
\begin{equation}
h_{vel}(\mathbf{q}, \mathbf{m}) = [\mathbf{m}^{T}(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{m}]^{-\frac{1}{2}} ( \mu \mathbf{n}^T \mathbf{m})
\end{equation}

$h_{EFORT}$ and $h_{vel}$ will favor contacts that allow large efforts or fast modifications in the velocity.
$EFORT$ in particular is useful for tasks such as standing up, pushing / pulling.
In other less demanding cases, manipulability can also be considered to avoid singularities.
To do so, we can consider the manipulability measure $h_{w}$, also given by \citeauthor{Yoshikawa1984}:

\begin{equation} \label{ellipsoid}
h_{w}(\mathbf{q}) = \sqrt{det(\mathbf{J}\mathbf{J}^T)}
\end{equation}
$h_{w}$ measures the ``distance'' of a given configuration to singularity. When $h_{w}$ is equal to 0, the configuration is singular;
the greater $h_{w}$ is, the further away the configuration is from singularity.
%~ We then derive:
%~ \begin{equation}
%~ h_{w}(\mathbf{q}, \mathbf{m}) = w(\mathbf{q})  (\mu \mathbf{n}^T \mathbf{m})
%~ \end{equation}

$h_{EFORT}$, $h_{vel}$ and $h_{w}$ define three kinematic heuristics, fast to compute (indeed, thanks to our sampling-based approach, the Jacobian and inverse products
can be precomputed off-line), that allow the planner to select the best
candidates according to user-defined criterion.
%~ This results in faster computation times, and a reduced number of overall contact switches.
